Improving the prospects for multimessenger astronomy with \earlywarning\ detection of compact binary coalescence.

### 2011-05-22 introduction.tex
added reference, editorial
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@758 ccdd05d5-56c1-461a-bf98-85f6876523b4

Prompt electromagnetic emission can arise as shells of relativistically
outflowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be the mechanism of short gamma-ray
bursts~\cite{nakar07}. This prompt emission is a probe into the extreme initial
conditions of the outflow. A very bright optical flash should accompany the
x-rays and gamma-rays~\cite{Sari99}. \editorial{Should mention that optical
flashes have been seen for a handful of long GRBs by very fast slews. TAROT has
seen many within ~20 seconds. \cite{Klotz09}} In contrast, an afterglow is
produced when the outflow impacts the interstellar medium.  Afterglows
represent lower Lorentz factors and the conditions of the environment far
outside of the central engine, and are relatively insensitive to initial
conditions. In order to observe the most intense electromagnetic emission and
provide an unbroken chain of observations between the progenitor system and the
expanding fireball, telescopes must point during the seconds surrounding the
merger.

\editorial{Citation needed for LOOC-UP} The gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
are observed.  In October 2010, \textsc{ligo} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\textsc{ligo} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration, namely \textsc{mbta}, ihope,
Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analysis
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current trigger generation and telescope actuation would then dominate
latency.

To this end, we have the ambition of reporting candidates not minutes
\emph{after} the merger, but seconds \emph{before}.  By looking for threshold
crossings before the signal leaves the detection band, it is possible to trade
some signal to noise ratio (\textsc{snr}) for latency.  Figure
\ref{fig:earlywarning} shows projected early trigger rates for NS--NS binaries
in Advanced \LIGO\ assuming the event rate predictions in
\cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_\mathrm{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources may be detected within 10
seconds of merger and $\sim1$ sources maybe detected within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above SNR 8.}
\end{figure}

>>>>

Prompt electromagnetic emission can arise as shells of relativistically
outflowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts~\cite{Lee:2005, nakar07}. This prompt emission is a probe into the
extreme initial conditions of the outflow. A very bright optical flash should
accompany the x-rays and gamma-rays~\cite{Sari99}. \editorial{Should mention
that optical flashes have been seen for a handful of long GRBs by very fast
slews. TAROT has seen many within ~20 seconds. \cite{Klotz09}} In contrast, an
afterglow is produced when the outflow impacts the interstellar medium.
Afterglows represent lower Lorentz factors and the conditions of the
environment far outside of the central engine, and are relatively insensitive
to initial conditions. In order to observe the most intense electromagnetic
emission and provide an unbroken chain of observations between the progenitor
system and the expanding fireball, telescopes must point during the seconds
surrounding the merger.

\editorial{Citation needed for LOOC-UP} The gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
are observed.  In October 2010, \textsc{ligo} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\textsc{ligo} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration, namely \textsc{mbta}, ihope,
Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current trigger generation and telescope actuation would then dominate
latency.

To this end, we have the ambition of reporting candidates not minutes
\emph{after} the merger, but seconds \emph{before}.  \editorial{I think we
should say also something about how we can also get the full SNR within moments
after the coalescence, it is obvious maybe, but we should be explicit}  By
looking for threshold crossings before the signal leaves the detection band, it
is possible to trade some signal to noise ratio (\textsc{snr}) for latency.
Figure \ref{fig:earlywarning} shows projected early trigger rates for NS--NS
binaries in Advanced \LIGO\ assuming the event rate predictions in
\cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_\mathrm{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources may be detected within 10
seconds of merger and $\sim1$ sources maybe detected within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above SNR 8.}
\end{figure}
